Article Title

### 2014-06-09 plos.tex
updated variance estimation

\section*{Introduction}
\emph{Functional connectivity} is a statistical description of observed \emph{multineuronal} activity patterns not reducible to the response properties of the individual cells. Functional connectivity reflects local synaptic connections, shared inputs from other regions, and endogenous network activity. Although functional connectivity is a phenomenological description without a strict mechanistic interpretation, it can be used to generate hypotheses about the anatomical architecture of the neural circuit and test hypotheses about the processing of information at the population level.

Pearson correlations between the spiking activity of pairs of neurons are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \ie the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect aspects of synaptic connectivity \cite{Gerstein:1964}.  Interest in neural correlations has been sustained by a series of discoveries of their nontrivial relationships to various aspects of circuit organization such as the physical distances between neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011, Ko:2013}, changes due to sensory stimulation and global brain states \cite{Greenberg:2008, Goard:2009, Kohn:2009, Rothschild:2010, Ecker:2010, Renart:2010}.

@@

To address these questions, we evaluated four regularized covariance matrix estimators that imposed different structures on the estimate. The estimators are designated as follows:
\begin{description}
\item[$C_{\sf sample}$] -- sample covariance matrix, the unbiased, unregularized estimator.
\item[$C_{\sf diag}$] -- linear shrinkage of covariances toward zero, \ie toward a diagonal covariance matrix.
\item[$C_{\sf factor}$] -- linear shrinkage of the covariances toward a low-rank matrix, representing inputs from unobserved factors (latent units).
\item[$C_{\sf sparse}$] -- sparse partial correlations, \ie a large fraction of the \emph{partial} correlations between pairs of neurons are set to zero.
\item[$C_{\sf sparse+latent}$] -- sparse partial correlations between the recorded neurons \emph{and} linear interactions with a number of latent units.
\end{description}

@@

We then performed a cross-validated evaluation to establish which of the four regularized estimators was most efficient for the population activity of dense groups of neurons in mouse primary visual cortex recorded with high-speed 3D random-access two-photon imaging of calcium signals. In our data, the sample correlation coefficients were largely positive and low.  We found that the best estimate of the correlation matrix was $C_{\sf sparse+latent}$.  This estimator revealed a sparse network of partial correlations, which we call \sq{interactions} here, between the observed neurons; it also inferred several latent units exerting linear effects on the observed neurons. We analyzed these networks of partial correlations and found the following: Whereas noise correlations were predominantly positive, the inferred interactions had a large fraction of negative values possibly reflecting inhibitory circuitry.  Moreover, we found that these interactions exhibited a stronger relationship to the physical distances and to the differences in preferred orientations than noise correlations. In contrast, the inferred negative interactions were less selective.

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.  The usual estimate of the covariance matrix is the \emph{sample covariance matrix} $C_{\sf sample}$ of the empirical sample of observations $x(t),\; t=1,\ldots,n$ as
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \frac 1 {\nu-1} \sum\limits_{t=1}^n (x(t)-\bar x)(x(t)-\bar x)^\T,\quad \bar x= \frac 1 n \sum\limits_{t=1}^n x(t)
\end{equation}
where $\nu$ denotes degrees of freedom in one neuron's activity within the sample ($\nu=n$ when observations are independent). In this study, we estimate the true mean, $\mu$, using the sample mean, $\bar x$, but seek a better estimate of the covariance matrix than the sample correlation matrix, $C_{\sf sample}$.

Given an covariance matrix estimate, the correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix.

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominantly in probablistic \emph{graphical modelling} wherein the joint distribution is explain by sets of two-way interactions \cite{Whittaker:1990}. Under the multivariate Gaussian distribution, zero partial correlations are equivalent to conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful measures of conditional independence and (lack of functinonal connectivity) between pairs of neurons.

@@

We considered four regularized estimators based on distinct families of reduced target estimates: $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$, and $C_{\sf sparse+latent}$. In probabilistic models with exclusively linear dependencies, the target estimates of these estimators correspond to distinct families of graphical models (\figref{1}{\,Row 1}).

@@

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive semidefinite matrix with low rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component: 
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (\figref{1}{Row 1, D}) \cite{Chandrasekaran:2010,Ma:2013}.

@@

The normal loss (Eq.~\ref{eq:loss}) is particularly convenient because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
\end{equation*}
With this property of additivity, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E[C_{\sf sample}^\prime]{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E[C_{\sf sample}^\prime]{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned to the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0066 to 0.043, mean 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals from eight cells out of 298 cells downsampled to 20 Hz. Spiking activity inferred by non-negative deconvolution is shown below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 298 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.017.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods for details). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.019, 0.0027, 0.0031, and 0.0022 nats/cell/bin, significantly greater than zero ($p<10^{-4}$ in each comparison, Wilcoxon signed rank test).

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
   {\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to the normal loss function (eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<10^{-4}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

Furthermore, estimator $C_{\sf sparse}$ outperformed estimator $C_{\sf factor}$ (Fig.~S\ref{supp:01}) suggesting the greater importance of modeling interactions between pairs of neurons than the common fluctuations of activity across the entire population.

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 89.9\% sparsity (or conversely, 11.1\% connectivity: $connectivity=1-sparsity$) with average node degree of 29.3 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 71, denoting 71 inferred latent units. The number of latent units and the average node degrees generally increased with population size, although these values varied widely between the datasets (\figref{5}{A and B}). There was an inverse relationship between the number of latent units and the average node degree (\figref{5}{C}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and needs further investigation.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure04.pdf}
    \end{center}
    \caption{{\bf Example of the structure revealed by $C_{\sf sparse+latent}$ the sparse+latent estimator.}
    {\bf A and B.} The regularized estimate $C_{\sf sparse+latent}$ closely approximates the sample correlation matrix $C_{\sf sample}$. 
    {\bf C and D.} However, the partial correlation matrices from the two estimates differ substantially.
    {\bf E.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 89.9\% off-diagonal zeros (bottom-left) and low-rank component of rank 71 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations: the gray interval indicates the range of correlations containing 89.9\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. The significant correlations were outside this interval. Yet 51.3\% of the interactions inferred by $C_{\sf sparse+latent}$  linked pairs of neurons whose correlation was below the threshold. Only the remaining 48.7\% of the interactions overlapped with sample correlations above the threshold.
    {\bf G.} A graphical depiction of the positive (green) and negative (magenta) partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the regularized partial correlations.
    {\bf I.} The same subset with sample correlations thresholded to match the sparsity of the regularized interactions.
}
\label{fig:4}
\end{FPfigure}

@@

When the cross-validation loss was not required, only the inner loop of cross-validation was used, applied to the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess-loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the sparse+latent estimator (Fig.~\ref{fig:4}--\ref{fig:6}).
\paragraph{Covariance estimation}
Within the inner loop of cross-validation, covariance matrix estimation was performed with fixed hyperparameter values provided by the search algorithm.  The computation of regularized estimators only required the sample covariance matrix $C_{\sf sample}$ of the training dataset.

>>>>

\section*{Introduction}
\emph{Functional connectivity} is a statistical description of observed \emph{multineuronal} activity patterns not reducible to the response properties of the individual cells. Functional connectivity reflects local synaptic connections, shared inputs from other regions, and endogenous network activity. Although functional connectivity is a phenomenological description without a strict mechanistic interpretation, it can be used to generate hypotheses about the anatomical architecture of the neural circuit and to test hypotheses about the processing of information at the population level.

Pearson correlations between the spiking activity of pairs of neurons are among the most familiar measures of functional connectivity \cite{Averbeck:2006, Zohary:1994, Kohn:2005, Bair:2001, Ecker:2010}.  In particular, \emph{noise correlations}, \ie the correlations of trial-to-trial response variability between pairs of neurons, have a profound impact on stimulus coding \cite{Zohary:1994, Abbott:1999, Sompolinsky:2001, Nirenberg:2003, Averbeck:2006, Josic:2009, Berens:2011, Ecker:2011}. In addition, noise correlations and correlations in spontaneous activity have been hypothesized to reflect aspects of synaptic connectivity \cite{Gerstein:1964}.  Interest in neural correlations has been sustained by a series of discoveries of their nontrivial relationships to various aspects of circuit organization such as the physical distances between the neurons \cite{Smith:2008, Denman:2013}, their synaptic connectivity \cite{Ko:2011},  stimulus response similarity \cite{Bair:2001, Arieli:1995, Chiu:2002, Kenet:2003, Kohn:2005, Cohen:2008, Cohen:2009, Ecker:2010, Rothschild:2010, Ko:2011, Smith:2013b}, cell types \cite{Hofer:2011}, cortical layer specificity \cite{Hansen:2012, Smith:2013}, progressive changes in development and in learning \cite{Golshani:2009, Gu:2011, Ko:2013}, changes due to sensory stimulation and global brain states \cite{Greenberg:2008, Goard:2009, Kohn:2009, Rothschild:2010, Ecker:2010, Renart:2010}.

@@

To address these questions, we evaluated four regularized covariance matrix estimators that imposed different structures on the estimate. The estimators are designated as follows:
\begin{description}
\item[$C_{\sf sample}$] -- sample covariance matrix, the unbiased, unregularized estimator.
\item[$C_{\sf diag}$] -- linear shrinkage of covariances toward zero, \ie toward a diagonal covariance matrix.
\item[$C_{\sf factor}$] -- a low-rank approximation of the sample covariance matrix, representing inputs from unobserved shared factors (latent units).
\item[$C_{\sf sparse}$] -- sparse partial correlations, \ie a large fraction of the \emph{partial} correlations between pairs of neurons are set to zero.
\item[$C_{\sf sparse+latent}$] -- sparse partial correlations between the recorded neurons \emph{and} linear interactions with a number of latent units.
\end{description}

@@

We then performed a cross-validated evaluation to establish which of the four regularized estimators was most efficient for the population activity of dense groups of neurons in mouse primary visual cortex recorded with high-speed 3D random-access two-photon imaging of calcium signals. In our data, the sample correlation coefficients were largely positive and low.  We found that the best estimate of the correlation matrix was $C_{\sf sparse+latent}$.  This estimator revealed a sparse network of partial correlations (\sq{interactions}), between the observed neurons; it also inferred several latent units exerting linear effects on the observed neurons. We analyzed these networks of partial correlations and found the following: Whereas significant noise correlations were predominantly positive, the inferred interactions had a large fraction of negative values possibly reflecting inhibitory circuitry.  Moreover, we found that these interactions exhibited a stronger relationship to the physical distances and to the differences in preferred orientations than noise correlations. In contrast, the inferred negative interactions were less selective.

\section*{Results}
% Results and Discussion can be combined.
\paragraph{Covariance estimation}
The covariance matrix is defined as
\begin{equation}\label{eq:true-covariance}
    \Sigma = \E{(x-\mu)(x-\mu)^\T},\quad \mu = \E{x}
\end{equation}
where $\E{\cdot}$ denotes expectation; the $p\times 1$ vector $x$ is a single observation of the firing rates of $p$ neurons over time $\Delta t$; and $\mu$ is the vector of expected firing rates.

Given a set of observations $\{x(t): t\in T$\} of population activity, where $x(t)$ is an $p\times 1$ vector of firing rates in time bin $t$, and an independent unbiased estimate $\bar x$ of the mean activity, the \emph{sample covariance matrix},
\begin{equation}\label{eq:sample}
	C_{\sf sample} = \left[\mbox{cov}\left(x(t)\right)\right]_{t\in T} =\frac 1 n \sum\limits_{t\in T} (x(t)-\bar x)(x(t)-\bar x)^\T,
\end{equation}
where $n$ is the size of $T$, yields an unbiased estimate of the covariance matrix so that $\E{C_{\sf sample}}=\Sigma$.

Given any covariance matrix estimate $C$, the corresponding correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix.

\emph{Pairwise partial correlations} are less familiar in the neuroscience community. 
The partial correlation between a pair of variables is the Pearson correlation coefficient of the residuals of the linear least-squares predictor of their activity based on all the other variables, excluding the pair \cite{Anderson:2003, Whittaker:1990}. Partial correlations figure prominently in probablistic \emph{graphical modelling} wherein the joint distribution is explained by sets of two-way interactions \cite{Whittaker:1990}. For the multivariate Gaussian distribution, zero partial correlations indicate conditional independence of the pair, implying a lack of direct interaction between them \cite{Dempster:1972, Whittaker:1990}. In the general case, partial correlations can serve as a measure of conditional independence under the assumption that most dependencies in the system include strong linear effects \cite{Whittaker:1990,Baba:2004}. As neural recordings become increasingly dense, partial correlations may prove useful as indicators of conditional independence (lack of functinonal connectivity) between pairs of neurons.

@@

We considered four regularized estimators based on distinct families of target estimates: $C_{\sf diag}$, $C_{\sf factor}$, $C_{\sf sparse}$, and $C_{\sf sparse+latent}$. In probabilistic models with exclusively linear dependencies, the target estimates of these estimators correspond to distinct families of graphical models (\figref{1}{\,Row 1}).

@@

Estimator $C_{\sf factor}$ approximates the covariance matrix by the factor model $L + D$, where $L$ is the $p\times p$ positive semidefinite matrix with low-rank and $D$ is a diagonal matrix of independent variances. The estimator, 
\begin{equation}\label{eq:c-factor}
C_{\sf factor} = L + (1-\lambda)D + \lambda \bar D,
\end{equation}
has two hyperparameters: the rank of $L$ (\ie the number of latent factors) and the shrinkage intensity $\lambda$. The structure imposed by $C_{\sf factor}$ favors conditions in which the population activity is linearly driven by a number of latent factors that affect many cells while direct interactions between the recorded cells are insignificant (\figref{1}{Row 1, B}).

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component: 
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low-rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (\figref{1}{Row 1, D}) \cite{Chandrasekaran:2010,Ma:2013}.

@@

The normal loss (Eq.~\ref{eq:loss}) is particularly convenient because it is additive in its second argument, \ie
\begin{equation*}\label{eq:additivity}
    \loss{C,z_1} + \loss{C,z_2} \equiv \loss{C,z_1+z_2}
\end{equation*}
Then, once $C$ is fixed, the validation loss is an unbiased estimate of the loss function:
\begin{equation*}
    \E{\loss{C,C_{\sf sample}^\prime}}=\loss{C,\E{C_{\sf sample}^\prime}}=\loss{C,\Sigma}.
\end{equation*}

@@

\paragraph{The $C_{\sf sparse+latent}$ estimator is most efficient in neural data}
We recorded the calcium activity of dense populations of neurons in the supragranular layers in primary visual cortex of fentanyl-anesthetized mice using fast random-access 3D scanning two-photon microscopy during visual stimulation (\figref{2}{A--B}) \cite{Reddy:2005, Katona:2012, Cotton:2013}. This technique allowed fast sampling (100--150 Hz) from large numbers (150--350) of cells in a small volume of cortical tissue ($200\times200\times100$ $\mu$m$^3$) in layers 2/3 and 4 (\figref{2}{C}).  The firing rates were inferred using sparse nonnegative deconvolution \cite{Vogelstein:2010} (\figref{2}{C}). Only cells that produced detectable calcium activity were included in the analysis (For details, see Methods).  First, 30 repetitions of full-field drifting gratings of 16 directions were presented in random order.  Each grating was presented for 500 ms, without intervening blanks.  This stimulus was used to compute the orientation tuning of the recorded cells (\figref{2}{D}). To estimate the noise correlation matrix, we presented only 2 distinct directions in some experiments or 5 directions in others with 100--300 repetitions of each direction. Each grating lasted 1 second and was followed by a 1-second blank.  The traces were then binned into 150 ms intervals aligned to the stimulus onset for the estimation of the covariance matrix.   The sample correlation coefficients were largely positive and low. The average value of the correlation coefficient across sites ranged from 0.0065 to 0.051, mean 0.018 (\figref{2}{F} and \figref{5}{D}).

\begin{figure}    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,center},capbesidewidth=8.3cm}}]{figure}[\FBwidth]
    {\caption{{\bf Acquisition of neural signals for the estimation of noise correlations.}
    Visual stimuli comprising full-field drifting gratings interleaved with blank screens ({\bf A}) presented during two-photon recordings of somatic calcium signals using fast 3D random-access microscopy ({\bf B}). 
    {\bf C--F.} Calcium activity data from an example site.
    {\bf C.} Representative calcium signals of seven cells, downsampled to 20 Hz, out of the 292 total recorded cells. Spiking activity inferred by non-negative deconvolution is shown by red ticks below the trace.
    {\bf D.} The spatial arrangement and orientation tuning of the 292 cells from the imaged site. Only spiking cells are shown. The cells' colors indicate their orientation preferences. The gray cells were not significantly tuned.
    {\bf E.} The sample noise correlation matrix of the activity of the neural population. 
    {\bf F.} Histogram of noise correlation coefficients. The red line indicates the mean correlation coefficient of 0.020.
} \label{fig:2}}
    {\includegraphics[width=8.3cm]{figures/Figure02.pdf}}
\end{figure}

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Methods for details). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<10^{-2}$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

\begin{figure}[!ht]
    \begin{center}
    \includegraphics{./figures/src/Fig3.eps}
    \end{center}
    \caption{
   {\bf Performance of estimator $C_{\sf sparse+latent}$ with respect to the normal loss function (eq.~\ref{eq:loss}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} 
    Covariance estimators $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ produced consistently greater validation losses than $C_{\sf sparse+latent}$ ($p<10^{-2}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). The box plots indicate the $25^{th}$, $50^{th}$, and $75^{th}$ percentiles with the whiskers extending to the minimum and maximum values after excluding the outliers marked with `+'. 
} \label{fig:3}
\end{figure}

@@

In the example site (Fig.~\ref{fig:4}), the sparse component had 92.8\% sparsity (or conversely, 7.2\% connectivity: $connectivity=1-sparsity$) with average node degree of 20.9 (\figref{4}{G}). The average node degree, \ie the average number of interactions linking each neuron, is related to connectivity as $degree = connectivity\cdot(p-1)$, where $p$ is the number of neurons. The low-rank component had rank 72, denoting 72 inferred latent units. The number of latent units and the average node degrees generally increased with population size, although these values varied widely between the datasets (\figref{5}{A and B}). There was an inverse relationship between the number of latent units and the average node degree (\figref{5}{C}): Several sites, despite their large population sizes, were driven by latent units and had few pairwise interactions. This variability may be explained by differences in brain states and recording quality and needs further investigation.

@@

\begin{FPfigure}
    \begin{center}
        \includegraphics[width=17.35cm]{./figures/Figure04.pdf}
    \end{center}
    \caption{{\bf Example of the structure revealed by $C_{\sf sparse+latent}$ the sparse+latent estimator.}
    {\bf A and B.} The regularized estimate $C_{\sf sparse+latent}$ closely approximates the sample correlation matrix $C_{\sf sample}$. 
    {\bf C and D.} However, the partial correlation matrices from the two estimates differ substantially.
    {\bf E.} The partial correlation matrix of the regularized estimate is decomposed into a sparse component with 92.8\% off-diagonal zeros (bottom-left) and low-rank component of rank 72 (top-right).
    {\bf F.} The sparse component of the regularized partial correlation matrix had little resemblance to the sample correlations: the gray interval indicates the range of correlations containing 92.8\% of cells pairs, equal to the fraction of zeros in the sparse partial correlation matrix. The significant correlations were outside this interval. Yet 72.3\% of the interactions inferred by $C_{\sf sparse+latent}$  linked pairs of neurons whose correlation was below the threshold. Only the remaining 27.7\% of the interactions overlapped with sample correlations above the threshold.
    {\bf G.} A graphical depiction of the positive (green) and negative (magenta) partial correlations as edges between observed neurons. The line density is proportional to the magnitude of the correlation.
    {\bf H.} A subset of neurons from the center of the cluster shown in {\bf G} showing the regularized partial correlations.
    {\bf I.} The same subset with sample correlations thresholded to match the sparsity of the regularized interactions.
}
\label{fig:4}
\end{FPfigure}

@@

When the cross-validation loss was not required, only the inner loop of cross-validation was used, applied to the entire dataset.  This approach was used to compute the covariance matrix estimates and their excess-loss in the simulation study (\figref{1}{\,Rows 4 and 5}) and to analyze the partial correlation structure of the sparse+latent estimator (Fig.~\ref{fig:4}--\ref{fig:6}).
\subsection*{Cross-validation with condition-specific variances}
The loss function defined in the main text assumes that a single covariance matrix is to be estimated. However, neuroscientists are commonly interested in estimating a single noise correlation matrix across multiple stimulus conditions when the response means and variances depend on the stimulus \cite{Vogels:1989, Ponce:2013}.

Perhaps the most obvious approach is to minimize a loss function that directly compares the regularized estimate of the correlation matrix to the sample correlation matrix of the validation dataset. Unfortunately, the sample correlation coefficient is not generally unbiased \cite{Fisher:1921}. Reducing the discrepancy from a biased estimate is not evidence of approaching the true value of the correlation matrix. Instead, we adapted the loss function (Eq.~\ref{eq:loss}) to the case where a single estimate of the correlation matrix is combined with condition-specific variance estimates.

The definitions provided in the main text suit the case when all observations are made in the same condition with the same mean and variances across all observations.  In this appendix, we extend these definitions to the case when a common correlation matrix needs to be estimated from measurements collected under several stimulus conditions with potentially different means and variances.

Let $\{x(t):t\in T_c\}$ be a set of measurements $x(t), t\in T_c$ collected in condition $c$. 
Here $n_c$ denotes the number of measurements in condition $c$, \ie the cardinality of $T_c$.

Similar to Eq.~\ref{eq:sample}, the sample covariance matrix for condition $c$ is 
\begin{equation}
	C_c = \left[\mbox{cov}(x(t))\right]_{t\in T_c}
	\equiv \frac 1{n_c} \sum\limits_{t\in T_c}\left(x(t)-\bar x_c\right)\left(x(t)-\bar x_c\right)^\T,
\end{equation}

Then the sample correlation matrix in condition $c$ is 
\begin{equation}
	R_c = V_c^{-\frac 1 2}C_cV_c^{-\frac 1 2},
\end{equation}
where $V_c = C_c \circ I$ is the diagonal matrix containing the sample variances.

Let $u_c = \frac{n_c}{\sum\limits_c n_c}$ denote the fraction of measurements in condition $c$.
Then the common correlation matrix $\bar R$ is estimated by averaging the condition-specific correlations:
\begin{equation}
	\bar R = \sum\limits_c u_c R_c 
	= \sum\limits_c u_c V_c^{-\frac 1 2} C_c V_c^{-\frac 1 2}
	= \sum\limits_c u_c \frac 1 {n_c} \sum\limits_{t \in T_c} z(t)z(t)^\T
	= \frac 1 n \sum\limits_{t} z(t)z(t)^\T
	= \mbox{cov}(z(t))
\end{equation}
where $z(t) = V_c^{\frac 1 2} \left(x(t) - \bar x_c\right)$ can be recognized as the z-score signal: signal $x(t)$ normalized by the the sample standard deviation specific to the condition. Then $\bar R$ is simply the covariance matrix of $z(t)$ across all conditions. This  method for computing neural correlations is most familiar to neuroscientists.

Although we are primarily interested in improved estimation of correlations, the loss function can be improved by improving the estimates of the condition-specific means and variances as well. Let $\hat V_c$ and $\hat R$ denote the regularized versions of $V_c$ and $\bar R$.

Then the regularized condition-specific covariance matrices are reconstituted as
\begin{equation}
	\hat C_c = \hat V_c^{\frac 1 2} \hat R \hat V_c^{\frac 1 2} 
\end{equation}

To evaluate the regularized estimate $\hat C_c$, independent sets of measurements $T_c^\prime$ of cardinality $n_c^\prime$ are reserved for the computation the condition-specific testing sample covariance matrices $C_c^\prime$:
\begin{equation}
	C_c^\prime = \frac 1{n_c^\prime}\sum\limits_{t\in T_c}\left(x(t)-\bar x_c\right)\left(x(t)-\bar x_c\right)^\T
\end{equation}
Note that the means $\bar x_c$ are carried over from the training dataset and not re-estimated from the testing data. This obviates the need to adjust the estimate for degrees of freedom.
